Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-25 paper.qmd
Updated Figure 14

<!--Fit the best model and compute error-->
```{r}
#| label: num-bins-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 15
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

# tr_from_to_df_pbmc <- tr_from_to_df_pbmc |>
#   filter(row_number() != 360)

@@

@@

trimesh_removed_pbmc <- vis_rmlg_mesh(
  distance_edges = distance_pbmc,
  benchmark_value = benchmark_pbmc,
  tr_coord_df = tr_from_to_df_pbmc,
  distance_col = "distance") +
  geom_point(
    data = tsne_pbmc_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    alpha = 0.2,
    color = "#000000"
  ) 
  #xlim(sc_xlims) + ylim(sc_ylims) +
  #interior_annotation("a", sc_ltr_pos)

@@

```{r}
#| label: proj-tsne-pbmc

## Hexagonal binning to have regular hexagons
hb_obj_pbmc <- hex_binning(
  data = tsne_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  r2 = r2_pbmc,
  q = 0.1)

tsne_data_with_hb_id <- hb_obj_pbmc$data_hb_id

df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID),
                                tsne_data_with_hb_id)

### Define type column
df <- df_all_pbmc |>
  dplyr::select(tidyselect::starts_with("PC")) |>
  dplyr::mutate(type = "data") ## original dataset

df_b <- df_bin_pbmc |>
  dplyr::filter(hb_id %in% df_bin_centroids_pbmc$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_pbmc$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)

df_exe <- dplyr::bind_rows(df_b, df)

## Set the maximum difference as the criteria
distance_df_small_edges <- distance_pbmc |>
  dplyr::filter(distance < benchmark_pbmc)
## Since erase brushing is considerd.

## First projection
projection <- cbind(
  c(0.038104,-0.001148,0.003905,-0.003332,-0.011747,-0.002688,0.003468,-0.014321,-0.000320),
  c(-0.000902,0.041414,0.005089,-0.000149,-0.000551,0.000239,-0.008013,-0.005799,-0.001411))

proj1_pbmc_tsne <- gen_proj_langevitour(
  points_df = df_exe,
  projection = projection,
  edge_df = distance_df_small_edges |> select(-distance)
)

## Second projection
projection <- cbind(
  c(0.013182,-0.022240,-0.001729,0.026946,-0.005289,0.001235,-0.007784,0.017500,-0.006949),
  c(0.027138,0.011939,0.009374,-0.004582,-0.004771,-0.027107,-0.005287,-0.002325,-0.007951))

proj2_pbmc_tsne <- gen_proj_langevitour(
  points_df = df_exe,
  projection = projection,
  edge_df = distance_df_small_edges |> select(-distance)
)

## Third projection
projection <- cbind(
  c(-0.003540,0.025008,0.023261,-0.012782,-0.011617,-0.006796,0.009566,-0.015022,0.001164),
  c(0.038201,-0.007751,0.005064,-0.014143,0.003255,-0.001091,0.001611,-0.003674,-0.008314))

proj3_pbmc_tsne <- gen_proj_langevitour(
  points_df = df_exe,
  projection = projection,
  edge_df = distance_df_small_edges |> select(-distance)
)

@@

```{r}
#| echo: false
#| fig-cap: "Model in $2\\text{-}D$, on the tSNE layout, and three views of the fit in projections from $9\\text{-}D$, for the PBMC3k data ($(s_1, \ s_2) = (-0.050, \ -0.058)$, $b = 300 \  (15, \ 20)$, $m = 136$, benchmark value to remove large edges is $0.133$). (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/5Y1hE4i7N2k>)."
#| label: fig-pbmc2-sc
#| fig-pos: H
#| out-height: 100%
#| eval: true

trimesh_removed_pbmc + proj1_pbmc_tsne + 
  proj2_pbmc_tsne + proj3_pbmc_tsne +
  plot_layout(guides='collect', ncol=2) &
  theme(legend.position='none')
```

>>>>

<!--Fit the best model and compute error-->
```{r}
#| label: num-bins-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 13
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

@@

# tr_from_to_df_pbmc <- tr_from_to_df_pbmc |>
#   filter(!(row_number() %in% c(65, 112, 130)))

@@

benchmark_pbmc <- 0.1

@@

tr_df <- distinct(tibble::tibble(
  x = c(tr_from_to_df_pbmc[["x_from"]], tr_from_to_df_pbmc[["x_to"]]), 
  y = c(tr_from_to_df_pbmc[["y_from"]], tr_from_to_df_pbmc[["y_to"]])))

tr_from_to_df_pbmc <- tr_from_to_df_pbmc |>
  filter(!(row_number() %in% c(155)))

trimesh_removed_pbmc <- ggplot() + 
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#33a02c") +
  geom_point(
    data = tsne_pbmc_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    alpha = 0.2,
    color = "#000000"
  )  +
  coord_fixed() +
  interior_annotation("a")

# trimesh_removed_pbmc <- vis_rmlg_mesh(
#   distance_edges = distance_pbmc,
#   benchmark_value = benchmark_pbmc,
#   tr_coord_df = tr_from_to_df_pbmc,
#   distance_col = "distance") +
#   geom_point(
#     data = tsne_pbmc_scaled,
#     aes(
#       x = tSNE1,
#       y = tSNE2
#     ),
#     alpha = 0.2,
#     color = "#000000"
#   ) 
#   #xlim(sc_xlims) + ylim(sc_ylims) +
#   #interior_annotation("a", sc_ltr_pos)

@@

```{r}
#| label: pbmc-tsne-model-proj

## Removed long edge in high-d
distance_df_small_edges_pbmc <- distance_df_small_edges_pbmc |>
  filter(!(row_number() %in% c(155)))

## First projection
projection <- cbind(
    c(0.40456,-0.55243,-0.58434,0.05063,-0.20540,-0.10469,0.28850,0.12209,0.17017),
    c(0.34225,0.53615,-0.05822,-0.01993,-0.55480,0.21337,0.34100,-0.07280,-0.33139))

pbmc_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("b")

## Second projection
projection <- cbind(
    c(0.07154,-0.44483,-0.04593,0.63343,0.21449,-0.39592,-0.29762,0.01726,0.30845),
    c(0.87788,-0.21837,0.11780,-0.17226,-0.17920,-0.00829,0.11424,0.28681,0.06091))

pbmc_proj_umap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("c")

## Third projection
projection <- cbind(
    c(0.13399,-0.00303,0.34841,-0.15626,-0.27432,0.48771,0.14252,0.39963,0.57983),
    c(0.49615,0.73920,-0.15299,-0.07463,-0.07948,0.13320,-0.15484,0.20042,-0.28869))

pbmc_proj_umap_model3 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("d")

@@

```{r}
#| echo: false
#| fig-cap: "Model in $2\\text{-}D$, on the tSNE layout, and three views of the fit in projections from $9\\text{-}D$, for the PBMC3k data ($(s_1, \ s_2) = (-0.050, \ -0.058)$, $b = 300 \  (15, \ 20)$, $m = 136$, benchmark value to remove large edges is $0.133$). (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/5Y1hE4i7N2k>)."
#| label: fig-pbmc2-sc
#| fig-pos: H
#| fig-width: 10
#| fig-height: 10

trimesh_removed_pbmc + pbmc_proj_umap_model1 + 
  pbmc_proj_umap_model2 + pbmc_proj_umap_model3 +
  plot_layout(guides='collect', ncol=2) &
  theme(legend.position='none')
```
